---
title: "A Quick Start Guide for wsrf"
author: "He Zhao, Graham Williams"
date: "`r Sys.Date()`"
bibliography:
  ./wsrf-guide.bib
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{A Quick Start Guide for wsrf}
output:
  knitr:::html_vignette:
    toc: yes
---

## Introduction ##

The [**wsrf**](http://cran.r-project.org/package=wsrf) package is a
parallel implementation of the Weighted Subspace Random Forest
algorithm (wsrf) of @xu2012classifying.  A novel variable weighting
method is used for variable subspace selection in place of the
traditional approach of random variable sampling.  This new approach
is particularly useful in building models for high dimensional data
--- often consisting of thousands of variables.  Parallel computation
is used to take advantage of multi-core machines and clusters of
machines to build random forest models from high dimensional data with
reduced elapsed times.

## Requirements and Installation Notes ##

Currently, **wsrf** requires R (>= 3.0.0),
[**Rcpp**](http://cran.r-project.org/package=Rcpp) (>= 0.10.2)
[@dirk2011rcpp; @dirk2013seamless].  For the use of multi-threading, a
C++ compiler with [C++11](https://en.wikipedia.org/wiki/C%2B%2B11)
standard support of threads or the [*Boost*](http://www.boost.org/)
C++ library with version above 1.54 is required. The choice is
available at installation time depending on what is available to the
user.  To install the latest version of the package, from within R
run:

```{r eval=FALSE}
install.packages("wsrf")
```

By default, multi-threading functionality is not enabled, which can be
configured through the argument `configure.args=`.

```{r eval=FALSE}
install.packages("wsrf",
                 type="source",
				 configure.args="--enable-c11=yes")
```

We recommend using C++11 standard library for accessing multi-threaded
functionality, which will be our main focus for development in the
future.  Though support for compiling C++11 code in packages is not
fully supported in current release of R, it has been tested that it
can be compiled if the user has already installed the latest version
of [*GCC*](https://gcc.gnu.org/) and C++ standard library^[C++11
support is experimental in R-devel but not yet finished, see
[Daily News about R-devel on
2013-12-02](http://developer.r-project.org/blosxom.cgi/R-devel/NEWS/2013/12/02#n2013-12-02)
and
[MinGW-w64 Notes by Duncan
Murdoch](https://rawgit.com/kevinushey/RToolsToolchainUpdate/master/mingwnotes.html)].

Besides the default installation for C++11, we also provide another
implementations of **wsrf**, which implements parallelism using
*Boost*.

The choice of version to install is available at installation.  The
version without parallelism, as required when C++11 is not available
nor is *Boost*, and is the recommended and only possible choice for
Microsoft Windows platform with the current version of R (3.2.0) (the
same as the first installation method above):

```{r eval=FALSE}
install.packages("wsrf",
                 configure.args="--enable-c11=no")
```

Finally the version using *Boost* for multithreading can be installed
with the appropriate configuration options. This is suitable when the
version of C++ available does not support C++11.

```{r eval=FALSE}
install.packages("wsrf",
                 type="source",
                 configure.args="--with-boost-include=<Boost include path>
                                 --with-boost-lib=<Boost lib path>")
```

## Usage ##

This section demonstrates how to use **wsrf**, especially on a cluster
of machines.

The example uses a small dataset *weather* from
[**rattle**](http://cran.r-project.org/package=rattle) [@rattle].  See
the help page of **rattle** in R (`?weather`) for more details of
*weather*.  Below are the basic information of it.

```{r usage_load, message=FALSE}
library("rattle")
ds <- weather
dim(ds)
names(ds)
```

Before building the model we need to prepare the training dataset.
First we note the various roles played by the different variables,
including identifying the irrelevant variables.

```{r usage_prepare}
target <- "RainTomorrow"
ignore <- c("Date", "Location", "RISK_MM")
(vars <- setdiff(names(ds), ignore))
dim(ds[vars])
```

Next we deal with missing values, using `na.roughfix()` from
**randomForest** to take care of them.

```{r message=FALSE}
library("randomForest")
if (sum(is.na(ds[vars]))) ds[vars] <- na.roughfix(ds[vars])
ds[target] <- as.factor(ds[[target]])
(tt <- table(ds[target]))
```

We construct the formula that describes the model which will predict
the target based on all other variables.

```{r}
(form <- as.formula(paste(target, "~ .")))
```

Finally we create the randomly selected training and test datasets,
setting a seed so that the results can be exactly replicated.

```{r}
seed <- 42
set.seed(seed)
length(train <- sample(nrow(ds), 0.7*nrow(ds)))
length(test <- setdiff(seq_len(nrow(ds)), train))
```

The signature of the function to build a weighted random forest model
in **wsrf** is:

```{r eval=FALSE}
wsrf(formula, 
     data, 
     ntrees=500, 
     nvars=NULL,
     weights=TRUE, 
     parallel=TRUE)
```

We use the training dataset to build a random forest model.  All
parameters, except `formula` and `data`, use their default values:
`500` for `ntrees` --- the number of trees; `TRUE` for `weights` ---
weighted subspace random forest or random forest; `TRUE` for
`parallel` --- use multi-thread or other options, etc.

```{r usage_build_by_default, message=FALSE}
library("wsrf")
model.wsrf.1 <- wsrf(form, data=ds[train, vars])
print(model.wsrf.1)
print(model.wsrf.1, 1)  # Print tree 1.
```

Then, `predict` the classes of test data.

```{r usage_evaluate}
cl <- predict(model.wsrf.1, newdata=ds[test, vars], type="class")
actual <- ds[test, target]
(accuracy.wsrf <- sum(cl == actual, na.rm=TRUE)/length(actual))
```

Thus, we have built a model that is around `r round(100*accuracy.wsrf,
0)`% accurate on unseen testing data.

Using different random seed, we obtain another model.

```{r usage_build_another, message=FALSE}
# Here we build another model without weighting.
model.wsrf.2 <- wsrf(form, data=ds[train, vars], weights=FALSE)
print(model.wsrf.2)
```

We can also derive a subset of the forest from the model or a
combination of multiple forests.

```{r usage_subset_combine}
submodel.wsrf <- subset.wsrf(model.wsrf.1, 1:150)
print(submodel.wsrf)
bigmodel.wsrf <- combine.wsrf(model.wsrf.1, model.wsrf.2)
print(bigmodel.wsrf)
```

Next, we will specify building the model on a cluster of servers.

```{r usage_build_on_cluster, eval=FALSE}
servers <- paste0("node", 31:40)
model.wsrf.3 <- wsrf(form, data=ds[train, vars], parallel=servers)
```

All we need is a character verctor specifying the hostnames of which
nodes to use, or a named integer vector, whose values of the elements
give how many threads to use for model building, in other words, how
many trees built simultaneously.  More detail descriptions about
**wsrf** are presented in the
[manual](http://cran.r-project.org/web/packages/wsrf/wsrf.pdf).

## References ##
